---
title: "User guide to the `dearseq` R package"
author: "Marine Gauthier, Denis Agniel, Boris Hejblum"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{dearseqUserguide} 
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---
  
```{r knitrsetup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = TRUE)
knitr::knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(0, 0, 0, 0))  # no margin
})
```

# Overview of dearseq

The `dearseq` (Differential Expression Analysis that Robustly controls the FDR *Gauthier et al., 2019*) `R` package tests gene expression for significance. `dearseq` relies on variance component score test accounting for data heteroscedasticity through precision weights. It performs both gene-wise and gene set analyses, and can deal with various experimental designs such as multiple biological conditions, repeated or longitudinal measurements. For more details, check the TODO [BioRxiv](lien).


# How to use dearseq for a non-longitudinal gene-wise analysis ? 

## A real dataset

Tuberculosis (TB) is a disease caused by a bacterium called Mycobacterium tuberculosis. Bacteria typically infect the lungs, but they can also affect other parts of the body. Tuberculosis can remain in a quiescent state in the body. It is called latent tuberculosis infection (LTBI). It is an infection without clinical signs, bacteriological and radiological disease.

*Berry et al.* first identified a whole blood 393 transcript signature for active TB using microarray analysis. Then, *Singhania et al.* found a 373-genes signature of active tuberculosis using RNA-Seq, confirming microarray results, that discriminates active tuberculosis from latently infected and healthy individuals.

We sought to investigate how many of the 373 genes *Singhania et al.* found using edgeR might actually be false positives. As a quick example, we propose a partial re-analysis (see *Gauthier et al.* for the complete analysis) of the *Singhania et al.* dataset:  we proceed to the Differential Expression Analysis (DEA) of the Active TB group against the Control one, omitting LTBI/Control and ActiveTB/LTBI comparisons. We focused on the Berry London dataset.

It results in 54 patients of whom 21 are active TB patients, 21 are latent TB patients and 12 are healthy control particiants.

For more details, check the article from *Berry  et al., 2010* [here](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3492754/).

The data are publicly available on [GEO website](https://www.ncbi.nlm.nih.gov/geo/) with GEO access number 'GSE107995'. We will be using the `GEOquery` package to get the data files from GEO website (see appendix for more details on `GEOquery`)

3 inputs are required to run `dearseq`:

* The gene expression matrix
* The matrix/vector of covariates that needs to be adjusted
* The matrix/vector of variables to be tested


### Design data matrix

In this example, we need to import the supplementary file available: 

```{r GEOquery, message=FALSE}
if (!requireNamespace("GEOquery", quietly = TRUE)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
        install.packages("BiocManager")
    }
    BiocManager::install("GEOquery")
}
```

```{r import, message=FALSE, results='hide'}
library(GEOquery)
GSE107991_metadata <- getGEO("GSE107991", GSEMatrix = FALSE)

get_info <- function(i){
    name <- Meta(GSMList(GSE107991_metadata)[[i]])$source_name_ch1
    name <- gsub("Active_TB", "ActiveTB", name)
    name <- gsub("Test_set", "TestSet", name)
    unlist(strsplit(name, split="_"))
}

infos <- vapply(1:length(GSMList(GSE107991_metadata)), FUN=get_info,
                FUN.VALUE = c("a", "b", "c", "d"))
infos_df <- cbind.data.frame("SampleID" = names(GSMList(GSE107991_metadata)), t(infos))
rownames(infos_df) <- names(GSMList(GSE107991_metadata))
colnames(infos_df)[-1] <- c("Cohort", "Location", "Set", "Status")

group_London_TB <- which(infos_df$Status=="ActiveTB") # 21
group_London_LTBI <- which(infos_df$Status=="LTBI") # 21
group_London_Control <- which(infos_df$Status=="Control") # 12

group <- rep(NA,length(infos_df$Status[-group_London_LTBI]))
group[which(infos_df$Status[-group_London_LTBI]=="ActiveTB")] <- rep(1,length(which(infos_df$Status[-group_London_LTBI]=="ActiveTB")))
group[which(infos_df$Status[-group_London_LTBI]=="Control")] <- rep(0,length(which(infos_df$Status[-group_London_LTBI]=="Control")))

```

We excluded the LTBI patients to focus on the comparison between Active TB patients and Control patients.


### Gene expression matrix

This matrix contains the gene expression (in cells) for each gene (in rows) of 
each sample (in columns) gathered from RNA-seq measurements. The gene expression 
could already be normalized or not. In this last case, the gene expression is 
normalized into log(counts) per million. 

The gene expression matrix (called `London`) is including in the package `dearseq`. It was extracted from one of the GEO supplementary files (namely the "GSE107991_edgeR_normalized_Berry_London.xlsx" file). 

NB: The data is already normalized using `edgeR`.

```{r readxl, message=FALSE}
if (!requireNamespace("readxl", quietly = TRUE)) {
    install.packages("readxl")
}
```

```{r London-download, message=FALSE}
library(readxl)
GEOquery::getGEOSuppFiles('GSE107991', filter_regex="edgeR_normalized_Berry_London")
London <- readxl::read_excel("GSE107991/GSE107991_edgeR_normalized_Berry_London.xlsx")
genes <- London$Genes
London <- as.matrix(London[, -c(1:3)])
rownames(London) <- genes
```

We have:

 * in row `ENSGxxxxxxxxxxx` for each gene identifier (saved in `genes`)
 * in column `Berry_London_Samplex` for the name of each sample
 * each cell contains the normalized gene expression
 
The vector `genes` contains all the gene identifiers.


The entire matrix contains $14,150$ genes and $54$ samples.


## Variance component score test:  identifying differentially expressed (DE) genes

This function provides the result of variance component score test for each gene.
`dearseq` function requires:

* `exprmat`: a numeric matrix containing the raw RNA-seq counts or preprocessed expressions
* `covariates`: a numeric matrix containing the model covariates that needs to be adjusted. Usually, its first column is the intercept (full of 1s).
* `variables2test`: a numeric design matrix containing the variables of interest (with whom the expression association is tested)
* `which_test`: a character string indicating which method to use to approximate the variance component score test, either "permutation" or "asymptotic".
* `preprocessed`: a logical flag indicating whether the expression data have already been preprocessed (e.g. log2 transformed). Default is FALSE, in which case y is assumed to contain raw counts and is normalized into log(counts) per million.
* `doPlot`: a logical flag indicating whether the mean-variance plot should be drawn.

We use the permutation test due to the low sample size (=33).

```{r LR_ST}
res_dearseq <- dearseq::dear_seq(exprmat = London[,-group_London_LTBI], 
                                 covariates = matrix(rep(1, ncol(London[, -group_London_LTBI])), ncol=1),
                                 variables2test = matrix(group, ncol=1),
                                 which_test='asymptotic', doPlot = FALSE,
                                 preprocessed=TRUE)
```

Now `res_dearseq` is a list object containing, in addition to the parameters summary: 

- the raw p-values
- the adjusted p-values (after correction for multiple testing with the Benjamini-Hochberg procedure)

Following *Singhania et al.*, to be included in the signature a DE gene must have
had both: 

- an absolute log2(fold change) > 1
- an FDR adjusted p-value < 0.05

### log FC with edgeR

To guaranty comparable results with the original edgeR analysis, we computed the log fold-changes using the `edgeR` procedure from the raw data.

```{r London-raw, message=FALSE}
GEOquery::getGEOSuppFiles('GSE107991', filter_regex="GSE107991_Raw_counts_Berry_London.xlsx")
London_raw <- readxl::read_excel("GSE107991/GSE107991_Raw_counts_Berry_London.xlsx")
genes_raw <- London_raw$Genes
London_raw <- as.matrix(London_raw[, -c(1:3)])
rownames(London_raw) <- genes_raw
```


```{r edgeR, message=FALSE}
if (!requireNamespace("edgeR", quietly = TRUE)) {
    install.packages("edgeR")
}
```

```{r logFC, message=FALSE}
library(edgeR)
Y <- rep(NA,length(infos_df$Status))
Y[which(infos_df$Status == "ActiveTB")] <- rep(1, length(which(infos_df$Status == "ActiveTB")))
Y[which(infos_df$Status == "LTBI")] <- rep(0, length(which(infos_df$Status == "LTBI")))
Y[which(infos_df$Status == "Control")] <- rep(0, length(which(infos_df$Status == "Control")))

dgList <- DGEList(counts=London_raw, group=Y)
cpm <- cpm(dgList)
countCheck <- cpm > 2
keep <- which(rowSums(countCheck) >= 5)
dgList <- dgList[keep, ]
dgList <- calcNormFactors(dgList, method = "TMM")
dgList_group <- dgList[, -group_London_LTBI]
design <- model.matrix( ~ group)
dgList_group <- estimateGLMTrendedDisp(dgList_group, design = design) 
fit <- glmFit(dgList_group, design)
lrt <- glmLRT(fit)
```

`lrt` contains the log fold-changes for the $14,150$ genes.

### Results 

```{r DEgenes}
df <- data.frame(logFC=lrt$table$logFC,res_dearseq$pvals$adjPval) 
res <- rbind(subset(df, res_dearseq$pvals$adjPval<.05 & logFC>1),subset(df, res_dearseq$pvals$adjPval<.05 & logFC<(-1)))
genes_dearseq <- genes[as.numeric(rownames(res))]
```

After applying the filtering, `genes_dearseq` returns the DE gene identifiers.

```{r singhania-373, message=FALSE}
# Singhania et al. 373 genes signature
download.file("https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6008327/bin/41467_2018_4579_MOESM3_ESM.xlsx", destfile = "GSE107991/41467_2018_4579_MOESM3_ESM.xlsx")
TB373 <- as.data.frame(readxl::read_excel("GSE107991/41467_2018_4579_MOESM3_ESM.xlsx"))
genes_373 <- TB373[-c(1:2),1]

```


# How to use dearseq for a longitudinal gene-set analysis ? 

# References

Singhania A, Verma R, Graham CM, Lee J, Tran T, Richardson M, et al., (2018) A modular transcriptional signature identifies phenotypic heterogeneity of human tuberculosis infection. _Nature communications_ 9(1):2308.

Berry, M. P. et al., (2010) An interferon-inducible neutrophil-driven blood transcriptional signature in human tuberculosis. _Nature_ 466, 973–977.


# Appendix

## GEOquery package

In case the data you want to analyze is publicly available through [Gene Expression Omnibus (GEO)](https://www.ncbi.nlm.nih.gov/geo/), you can access it with the `GEOquery` package, that can be installed with the following commands:

```{r dl_GEOquery, warning=FALSE, message=FALSE, eval=FALSE}
if (!requireNamespace("GEOquery", quietly = TRUE)) {
    if (!requireNamespace("BiocManager", quietly = TRUE)){
        install.packages("BiocManager")
    }
    BiocManager::install("GEOquery")
}
```

More details can be found on [Bioconductor](https://bioconductor.org/packages/release/bioc/html/GEOquery.html) and in Davis S, Meltzer P, (2007) GEOquery: a bridge between the Gene Expression Omnibus (GEO) and Bioconductor _Bioinformatics_ 14:1846-1847.
